Predicting Conflict Escalation

### 2023-04-25 docs/hes_final/article.qmd
transformer section rewrite

Our second approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity we will refer to the first of these models as “the encoder” and the second as the “the classifier”.

Prior to training, the GDELT dataset was split into one 512xn matrix for each country and month where n is the number of articles in the dataset published in that month related to that country. Each of these matrices was then split into several random subsamples of 512x50. Samples for countries in which n<50 were padded, and multiple samples were taken for countries in which n >50. This resampling dramatically increased the number of training samples by acting as though the same month had occurred multiple times while still maintaining uniqueness of each sample. Simultaneously, limiting the size of each sample to 50 articles reduced the dimensionality of each sample, reducing the risk of overfitting.

The encoder was then trained using these samples, with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons, and its weights were frozen. The encoder was then used to scan each of the original 512xn matrices, resulting in a length-16 vector for each 512x50 subsample. This reduced the dimensions of each sample to 16x(n/50).

The classifier was then trained on this new, distilled dataset without resampling. The same time-shifted spike metric was again used as the training target.  The reduced dimensionality of this dataset was sufficient to offset the risk of overfitting.

>>>>

Our second approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.

The purpose of the encoder was to reduce the dimensionality of an embedding corpus in a manner analogous to a convolutional layer in a convolutional neural network. The classifier could then be used to make final predictions on a distilled dataset.

Prior to training, a base dataset was created by splitting the GDELT dataset into one $512xn$ matrix for each country and month where $n$ is the number of articles in the dataset published in that month related to that country. Each of these matrices was then split into several random subsamples of $512x50$. Samples for countries in which $n<50$ were padded. This process had two effects on training. First, by limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample. Simultaneously, by reusing the same training label for multiple unique training samples, we were able to treat each month as though it had occurred multiple times. This greatly increased the size of the training dataset and helped overcome the limitations imposed by the limited time window available in the GDELT embedding corpus.

This new dataset represents a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive less. Despite this limitation, this dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons.

With the encoder's training complete, we returned to the original base dataset. Each country-month matrix was again split into $512x50$ matrices, this time without resampling. Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector. Each vector was then stacked, resulting in a single matrix for each country and month of size $16x\frac{n}{50}$. This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.
